Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-06 paper.qmd
some very minor fixes, feel free to undo

The second example shows a longitudinal record of a single patient, measured at ages 45, 50, 55, 65 and 70. The lines connect the records in tome order. The projection corresponds to the maxima from a projection-pursuit guided tour using the anomaly index:

@@

Two projections of simulated example data corresponding to common liver tests: (a) sample of female patients, (b) longitudinal test results for a single patient. Red cross indicates observation is outide the 4-D confidence ellipse. The female patients tend to have lower ALP and ALT than the normal range. As the patient aged, the level of ALP increases and ALT decreases.
:::

@@

The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions, see Fig. @fig-weather (a). It provides a global picture showing where the outlying points differ from the normal distribution. We can see that the contrast between maximum (atmx) and minimum (atmn) temperature is primarily used in the direction where points are most outlying. In the orthogonal direction the variables precipitation (prc) and number of days with maximum temperature above 25°C (nsm) can be used to separate the clusters 2, 3 and 5. With this solution the two points in cluster 4 cannot be resolved, they fall inside the cluster. To better understand what is different for those two years we separately run the anomaly index for this subset, see Fig. @fig-weather (b). From the result we see that the maximum temperature is not relevant for this cluster, but that they have low values for the minimum temperature. We also see that there is still some averaging, and the preferred solution is not placing either point on the outside of the projected cluster.

@@

The methods on directional outlyingness or Stahel-Donoho outlyingness, described above, leng themselves to new potential \bD{} projection pursuit indexes. These would integrate nicely into a \bD{} projection pursuit guided tour to interactively visualise the potential outliers relative to the rest of the sample.

>>>>

The second example shows a longitudinal record of a single patient, measured at ages 45, 50, 55, 65 and 70. The lines connect the records in time order. The projection corresponds to the maxima from a projection-pursuit guided tour using the anomaly index:

@@

Two projections of simulated example data corresponding to common liver tests: (a) sample of female patients, (b) longitudinal test results for a single patient. Red cross indicates observation is outide the 4-D confidence ellipse. The female patients tend to have lower ALP and ALT than the normal range. As the patient ages, the level of ALP increases and ALT decreases.
:::

@@

The new anomaly index is first applied to the full dataset, such that the final projection will be affected by averaging distance of points in many directions, see Fig. @fig-weather (a). It provides a global picture showing where the outlying points differ from the normal distribution. We can see that the contrast between maximum (atmx) and minimum (atmn) temperature is primarily used in the direction where points are most outlying. In the orthogonal direction the variables precipitation (prc) and number of days with maximum temperature above 25°C (nsm) can be used to separate the clusters 2, 3 and 5. With this solution the two points in cluster 4 cannot be resolved, they fall inside the ellipse. To better understand what is different for those two years we separately run the anomaly index for this subset, see Fig. @fig-weather (b). From the result we see that the maximum temperature is not relevant for this cluster, but that they have low values for the minimum temperature. We also see that there is still some averaging, and the preferred solution is not placing either point on the outside of the projected ellipse.

@@

The methods on directional outlyingness or Stahel-Donoho outlyingness, described above, lend themselves to new potential \bD{} projection pursuit indexes. These would integrate nicely into a \bD{} projection pursuit guided tour to interactively visualise the potential outliers relative to the rest of the sample.
